|

### 2023-10-16 chapter_2/src/roberts_empirical_chapter_1.qmd
doing a bit more writing

---
title: |
  Does Color Convey Political Information?
short-title: Color as Political Information
published: "Manuscript draft: please do not share without author permission."
code-repo: "For replication, go to: <https:://github.com/DamonCharlesRoberts/book_project>."
author:
  - name: Damon C. Roberts
    email: damon.roberts-1@colorado.edu
    orcid: 0000-0002-4360-3675
    title: PhD Candidate
    affiliations:
      - id: CU
        name: University of Colorado Boulder
        department: Political Science
        address: 333 UCB
        city: Boulder
        region: CO 
        postal-code: 80309-0333
abstract: |
  Do the colors red and blue convey political information? If so, how do they shape candidate preferences and vote choice? Political psychologists rely on interdisciplinary approaches to construct theories of how we process political information. Likewise, those who study visual political information come from a number of disciplinary traditions to study how politics may be conveyed visually. As these groups of scholars are working in parallel, they have yet to fully integrate their work to construct a theory explaining the cognitive and behavioral consequences of visual information on political attitude expression and behavior. In this chapter, I present a snap-judgment model of visual information processing to demonstrate how visual information efficiently conveys substantive political information. Specifically, I focus on a simple but obvious case that we only have anecdotal and descriptive evidence of: the association between Republicans with the color red and Democrats with the color blue. First, I use supervised machine learning that does a pixel-by-pixel classification of the color on yard signs used by campaigns for the U.S. House of Representatives from 2016 to 2020 and merge this with election return data from the MIT Election Labs. From these data, I find that systematic descriptive evidence of campaigns using more red in their yard signs in districts that are safe for Republicans and campaigns using more blue in their yard signs in districts that are safe for Democrats. Next, I turn to an experiment to examine whether these associations of the colors red and blue with Republicans and Democrats exist among voters and whether it shapes attitude expression and behavior. I find that these associations of the color red with Republicans and the color blue with Democrats do exist among the public and that these associations are strong enough to shape support for political candidates. This chapter advances our understanding in a few areas: that visual information is a pre-cognitive form of information that we should integrate with our models of information processing that consider more traditional forms of political information, provides a cognitive mechanism to explain how the public have come to associate the colors red and blue with the two dominant political parties, as well as demonstrates that these cognitive processes are capable of influencing behavior as important as vote choice.
additional-info: |
  I have no known conflict of interest to disclose.
keywords:
  - visual information
  - information processing
  - partisanship
date: today
bibliography: "../../assets/candidate_support_chapter_references.bib"
format:
  hikmah-pdf:
    cite-method: citeproc
    latex-output-dir: ../out/
    geometry:
      - top=1in
      - bottom=1in
      - left=1in
      - right=1in
      - heightrounded
    linestretch: 2
    fontsize: 12pt
    appendix-style: default
    include-in-header:
      text: |
        \usepackage{pdflscape}
        \newcommand{\blandscape}{\begin{landscape}}
        \newcommand{\elandscape}{\end{landscape}}
linestretch: 2
execute:
  echo: false
  warning: false
  message: false
params:
  replicate: false
---

@@

Building upon descriptive evidence suggesting that Democratic campaigns use the color blue more than Republican campaigns [@williams_whats_2022] and that Republican voters express a preference for red during election season [@schloss_politics_2014], I present a snap-judgment model of politically-relevant visual information processing that outlines the cognitive processes that drive these associations and how those associations drive important political outcomes. I use theories of associative memory and spreading activation [see @collins_spreading-activation_1975] to conceptualize the association of the colors red and blue with Republicans and Democrats as two interconnected nodes. Informed by conceptualizations of these connections as carrying affective information informed by perspectives on biological survival reaching back to Darwin [@ralph_neuroscience_2018], I argue that when these nodes are activated by exposure to these colors in political contexts, other related nodes become active and contain affective information. As partisanship evokes strong affective reactions for many members of the public [@iyengar_affect_2012], this activation elicits an affective reaction. Considering that most visual information and that these processes of association occur pre-consciously [@ames_impression_2012;@fazio_2007;@sander_cambridge_2013], I argue that the elicited affective response to such information can be "transferred" [see @morris_activation_2003;@taber_illusion_2016] to later post-cognitive processing of more traditional forms of political information such as stated policy positions. The implication being that these associations can inform reactions to political information and thus political behavior subconsciously.
<!--
This chapter applies the snap-judgment model to a simple case: the use of color on yard signs used by campaigns in elections. I focus on yard signs as they are a common form of campaign marketing and evidence suggests that they can shape individual behavior [@makse_politics_2019] and electoral outcomes [@green_effects_2016]. Additionally, I start by examining the colors red and blue. While there are other colors that may provide political information -- black and white for Black Lives Matter and Blue Lives Matter movements, the overarching goal of this book is to lay the foundation for a larger research agenda in visual political psychology [@bucy_editors_2021]. Without such foundations, we would be jumping into the study of complex cases and circumstances without a firm grasp of how things matter on a more fundamental level.

@@

These findings bring much needed evidence in support of the conventional wisdom that the colors red and blue convey political information. Perhaps what is most intriguing about my results is that I find that the colors red and blue alone seem to make individuals willing to vote for a candidate. The association between the colors red and blue are strong enough that individuals express a willingness to cast a vote for a candidate based on the color of the candidate's yard sign; this occurs without the individual confirming that the candidate matches their policy stances or partisan affiliation. These tendencies, even if only based on a hunch among the professionals on the ground, seem to lead to descriptive differences in the use of these colors with respect to strategy. The following chapters dig a bit deeper into the idea that colors can shape attitudes, pitting the information individuals get from color against information gleaned from explicit policy stances.

@@

In line with existing theories of *pip*, I use a popular conceptualization of attitudes as emerging from accessed associated memories [see @lodge_rationalizing_2013; see also @collins_spreading-activation_1975 for a perspective in a different application]. The implication of such a conceptualization is that attitudes may be unstable (though not in a random sense). That is, as attitudes are associative, they manifest slightly differently depending on the associative paths activated [@fazio_attitudes_2007]. The retrieval of relevant memories to the attitude depends on many factors, such as the recency of the event, the similarity of the context, and the importance or salience of the memory [@kahana_laws_2022]. This means that the memories retrieved to contribute to an attitude that is quite variable. However, to understand where that variability comes from, we must understand the deeper processes influencing how information is encoded and later retrieved. This illustrates my need to build upon the reigning models of *pip*. Colors may act as a contextual feature that may lead to variability in how a set of political information may shape attitudes. Let me explain how colors might do this.

@@

What the snap-judgment model predicts is happening in your head is something like the following: as soon as the light bounces off the sign (to produce a particular wavelength) and reaches your eyes, your brain is already trying to make sense of this information. This is a valuable tool for survival that biology has optimized for millions of years [@parker_blink_2003]. Rather than slowly processing the visual information and finding yourself in the jaws of a predator, or processing it quickly but forming the wrong impression and running away from a friend, the brain processes the information quickly and pre-cognitively [@newell_unified_1990]. To make sense of such information, it accesses familiar information similar to what it is currently attempting to process, and it does this for the purposes of efficiency [@kahana_laws_2022]. This means accessing memories that contain valanced information [@kensinger_affective_2022]: should I avoid this, or is it pleasant? Once the brain has finished such processing, it can pass its prediction to your cognitive memory. Once you form a reflex of avoid or approach, this opens up space for your brain to process the more complex information: to take the patterns of the light as shapes that construct symbols and letters. This comes later because this information not only requires access to information about what it *is*, but also about what it *means*. Once you understand what it means, you have the information necessary to evaluate it.

@@

In [Study 1](#study-1) I start with testing $H_5$. In some informal conversations that I had with political campaign professionals, it became clear that the descriptive evidence connecting the colors red and blue with Republicans and Democrats is still very much based on experience.

@@

It appears that campaigns recognize that these are not simply patterns that journalists [see @elving_color_2014] and researchers [see @williams_whats_2022] have picked up on, bit that they are associations held by voters and campaigns respond accordingly:

>>>>

---
title: |
  Does Color Convey Political Information?
short-title: Color as Political Information
published: "Manuscript draft: please do not share without author permission."
code-repo: "For replication, go to: <https:://github.com/DamonCharlesRoberts/book_project>."
author:
  - name: Damon C. Roberts
    email: damon.roberts-1@colorado.edu
    orcid: 0000-0002-4360-3675
    title: PhD Candidate
    affiliations:
      - id: CU
        name: University of Colorado Boulder
        department: Political Science
        address: 333 UCB
        city: Boulder
        region: CO 
        postal-code: 80309-0333
abstract: |
  Do the colors red and blue convey political information? While decisions of political branding are assumed to be significant for electoral outcomes, we have few theories about the cognitive processes underlying the processing of such information and whether these decisions influence realized political behavior. Building upon interdisciplinary perspectives, this chapter introduces the snap-judgment model of visual political information processing. The theory argues that color allows individuals to form pre-cognitive attitudes of a political object and that these pre-cognitive attitudes shape latter and more cognitively laborious information processing and behavior. Meaning, that we can activate partisan attachments automatically through information as simple as color. In the first study, I talk to a political marketing firm and use a pixel-by-pixel classification of images from over 1000 yard signs for elections in the U.S. House of Representatives. Evidence suggests that in districts where Democrats have an electoral advantage, there is more blue on the yard signs in the district. To examine the cognitive processes, I use a novel experimental design by tracking cursor movements and limiting participant's ability to view a yard sign to only 5000ms. I find that participants do make these associations when features of the yard sign are held constant and find that Republicans prefer candidates with red yard signs and that Democrats prefer candidates with blue yard signs.
additional-info: |
  I have no known conflict of interest to disclose.
keywords:
  - visual information
  - information processing
  - partisanship
date: today
bibliography: "../../assets/candidate_support_chapter_references.bib"
format:
  hikmah-pdf:
    cite-method: citeproc
    latex-output-dir: ../out/
    geometry:
      - top=1in
      - bottom=1in
      - left=1in
      - right=1in
      - heightrounded
    linestretch: 2
    fontsize: 12pt
    appendix-style: default
    include-in-header:
      text: |
        \usepackage{pdflscape}
        \newcommand{\blandscape}{\begin{landscape}}
        \newcommand{\elandscape}{\end{landscape}}
linestretch: 2
execute:
  echo: false
  warning: false
  message: false
params:
  replicate: false
---

{{< pagebreak >}}

@@

Building upon descriptive evidence suggesting that Democratic campaigns use the color blue more than Republican campaigns [@williams_whats_2022] and that Republican voters express a preference for red during election season [@schloss_politics_2014], I present a snap-judgment model of politically-relevant visual information processing that outlines the cognitive processes that drive these associations and how those associations drive important political outcomes. I use theories of associative memory and spreading activation [see @collins_spreading-activation_1975] to conceptualize the association of the colors red and blue with Republicans and Democrats as two interconnected nodes. Informed by conceptualizations of these connections as carrying affective information informed by perspectives on biological survival reaching back to Darwin [@ralph_neuroscience_2018], I argue that when these nodes are activated by exposure to these colors in political contexts, other related nodes become active and contain affective information. As partisanship evokes strong affective reactions for many members of the public [@iyengar_affect_2012], this activation elicits an affective reaction. Considering that most visual information and that these processes of association occur pre-cognitively [@ames_impression_2012;@fazio_attitudes_2007;@sander_cambridge_2013], I argue that the elicited affective response to such information can be "transferred" [see @morris_activation_2003;@taber_illusion_2016] to later post-cognitive processing of more traditional forms of political information such as stated policy positions. The implication being that these associations can inform reactions to political information and thus political behavior pre-cognitively.
<!--
This chapter applies the snap-judgment model to a simple case: the use of color on yard signs used by campaigns in elections. I focus on yard signs as they are a common form of campaign marketing and evidence suggests that they can shape individual behavior [@makse_politics_2019] and electoral outcomes [@green_effects_2016]. Additionally, I start by examining the colors red and blue. While there are other colors that may provide political information -- black and white for Black Lives Matter and Blue Lives Matter movements, the overarching goal of this book is to lay the foundation for a larger research agenda in visual political psychology [@bucy_editors_2021]. Without such foundations, we would be jumping into the study of complex cases and circumstances without a firm grasp of how things matter on a more fundamental level.

@@

These findings bring much needed evidence in support of the conventional wisdom that the colors red and blue convey political information. The implication of the theory and these findings are that partisan attachments are automatically activated through things as seemingly simple and ubiquitous as color. These activated partisan attachments then have causal effects for attitude expression and political behavior. Given the existing literature demonstrating low-levels of political engagement among the mass public [see @hibbing_stealth_2002] as well as the current hyper-polarized era [see @iyengar_affect_2012], this chapter's argument and findings suggests that voters need not consume any substantive political information before expressing a willingness to cast a vote for a candidate.

@@

I use a popular conceptualization of attitudes as emerging from accessed associated memories [see @lodge_rationalizing_2013; see also @collins_spreading-activation_1975]. The implication of such a conceptualization is that attitudes may be unstable (though not in a random sense). That is, as attitudes are associative, they manifest slightly differently depending on the associative paths activated [@fazio_attitudes_2007]. The retrieval of relevant memories to the attitude depends on many factors, such as the recency of the event, the similarity of the context, and the importance or salience of the memory [@kahana_laws_2022]. This means that the memories retrieved to contribute to an attitude that is quite variable. However, to understand where that variability comes from, we must understand the deeper processes influencing how information is encoded and later retrieved. This illustrates my need to build upon the reigning models of *pip*. Colors may act as a contextual feature that may lead to variability in how a set of political information may shape attitudes. Let me explain how colors might do this.

@@

What the snap-judgment model predicts is happening in your head would be the following: as soon as the light bounces off the sign (to produce a particular wavelength) and reaches your eyes, your brain is already trying to make sense of this information. This is a valuable tool for survival that biology has optimized for millions of years [@parker_blink_2003]. Rather than slowly processing the visual information and finding yourself in the jaws of a predator, or processing it quickly but forming the wrong impression and running away from a friend, the brain processes the information quickly and pre-cognitively [@newell_unified_1990]. To make sense of such information, it accesses familiar information similar to what it is currently attempting to process, and it does this for the purposes of efficiency [@kahana_laws_2022]. This means accessing memories that contain valanced information [@kensinger_affective_2022]: should I avoid this, or is it pleasant? Once the brain has finished such processing, it can pass its prediction to your cognitive memory. Once you form a reflex of avoid or approach, this opens up space for your brain to process the more complex information: to take the patterns of the light as shapes that construct symbols and letters. This comes later because this information not only requires access to information about what it *is*, but also about what it *means*. Once you understand what it means, you have the information necessary to evaluate it.

@@

In [Study 1](#study-1) I start with testing $H_5$. Though my focus is on the cognitive mechanisms for individuals, I first wanted to examine whether practitioners of political marketing also perceive that these associations exist and are influential. In some informal conversations that I had with political campaign professionals, it became clear that the descriptive evidence connecting the colors red and blue with Republicans and Democrats is still very much based on instinct rather than empirical evidence.

@@

It appears that campaigns recognize that these are not simply patterns that journalists [see @elving_color_2014] and researchers [see @williams_whats_2022] have picked up on, but that they are associations held by voters and campaigns respond accordingly:
